\texttt{aeons}: approximating the end of nested sampling

### 2023-12-28 paper/paper.tex
correcting logLlivemax position
changes based on johannes' comments

\subsection{Log-likelihood}\label{sec:logL}
We now consider the distribution of the samples in log-likelihood. To get an insight into the analytics, we will examine the representative case of the $d$-dimensional multivariate Gaussian with maximum point $\logLmax$ and length scale $\sigma$:
\begin{equation}\label{eq:gaussian_logL}
    \log \Like = \logLmax - X^{2/d}/2\sigma^2.
\end{equation}
Providing that $\sigma^d \ll 1$, the posterior can be approximated as 
\begin{equation}
    \mathcal{P}(X) = \tfrac{1}{(2\pi\sigma^2)^{d/2}} e^{-X^{2/d}/2\sigma^2}, 
    \label{eqn:PX}
\end{equation}
which in terms of $\log\mathcal{L}$ becomes
\begin{equation}
    \mathcal{P}(\log\mathcal{L}) = \frac{1}{\Gamma\left(\frac{d}{2}\right)}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
\end{equation}
i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, with mean and variance
\begin{equation}
    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
Given the prior is uniform $X\sim U(0,1)$, the KL divergence in the $\sigma^d \ll 1$ limit is:
\begin{equation}
\mathcal{D}_\mathrm{KL} = -\frac{d}{2}\log 2 \Gamma(1+\tfrac{d}{2})^{2/d} \sigma^2
\end{equation}

@@

\subsection{The termination prior volume}
Making the above discussion more precise, we wish to find the compression factor $\log X_ \mathrm{f}$ at which the termination criterion is met, which is larger in magnitude than $\DKL$ (\cref{fig:logX_distribution}). The difficulty is that at an intermediate iteration we only know the posterior up to the maximum log-likelihood live point, which until just before the end is far from the posterior bulk. 
\par
In order to get an idea of where the true posterior bulk sits, we need to predict what the posterior looks like past the highest live point. We do this by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\Like(X)$ traced out by the live and dead points. 
One would never use this predicted posterior to perform inference, since more accuracy can always be achieved by simply finishing the run. However, we will demonstrate it is sufficient for making a run-time prediction for $\log X_\mathrm{f}$. 
\par
Quantitatively, this proceeds as follows: fit a function $f(X, \phi)$  with some parameters $\phi$ to the known likelihood profile, which allows us to express the prior volume we need to compress to as
\begin{equation}
	\Delta \mathcal{Z} = \epsilon \mathcal{Z}_\mathrm{tot},
\end{equation}
or equivalently
\begin{equation}\label{endpoint}
	\int_0^{X_\mathrm{f}} f(X, \phi)\ \mathrm{d}X = \epsilon \left( \int_0^{X_i} f(X, \phi)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead} \right),
\end{equation}
where $X_i$ is the volume of the iteration we have currently compressed to, and $\mathcal{Z}_\mathrm{dead}$ is the evidence we have accumulated up to this point. $X_\mathrm{f}$ can then be identified by solving the above equation either analytically or numerically. 
\par
Once $X_\mathrm{f}$ is known, the corresponding iteration count depends on the live point schedule. For example, in the constant $\nlive$ case $\log X$ decreases by $1/n$ at each iteration, so the total number of iterations $N_\mathrm{f}$ would be
\begin{equation}
	N_\mathrm{f} = - n\log X_\mathrm{f} .
\end{equation}

@@

\subsection{Alternative approaches}
More comprehensive Bayesian approaches, perhaps including \textit{a priori} information about the likelihood or greater flexibility in the fitting function, could likely perform better than what we have just presented. However, such methods would not befit \textit{run-time} prediction which has a much more limited computational budget, hence the more pragmatic approach we have adopted. Here, we discuss as a benchmark alternative approaches to endpoint estimation that have a comparable computational complexity. 
\subsubsection{Integral progress}
An alternative approach used in \textsc{Ultranest} \citep{ultranest} derives a progress bar based on the fraction of the accumulated integral compared to the remaining integral, approximated as 
\begin{equation}\label{eq:integral_progress}
    \mathcal{Z}_\mathrm{rem} \approx \Like_\mathrm{max}^{\mathrm{live}} X_i.
\end{equation}
This has a several shortcomings. First, run-time is proportional to compression rather than accumulation of the integral, since it takes just as long to traverse the width of the bulk as it does any other width. Second, because of the point-like nature of the posterior mid-run, the remaining integral approximated as such holds nearly all of the evidence, so the relative fraction of the accumulated and remaining evidence is almost zero for most of the run. Finally, approximation \eqref{eq:integral_progress} is always an underestimate, because as previously found the maximum live point is generally nowhere near the true maximum. This approach can however be useful in the low dimensions appropriate for \textsc{Ultranest} when the live points are always near the maximum, but in general is less reliable.

>>>>

\subsection{Log-likelihood}\label{sec:logL}
We now consider the distribution of the samples in log-likelihood. To get an insight into the analytics, we will examine the representative case of the $d$-dimensional multivariate Gaussian with maximum point $\logLmax$ and length scale $\sigma$:
\begin{equation}\label{eq:gaussian_logL}
    \log \Like = \logLmax - X^{2/d}/2\sigma^2.
\end{equation}
Providing that $\sigma^d \ll 1$, the posterior can be approximated as 
\begin{equation}
    \mathcal{P}(X) = \tfrac{1}{(2\pi\sigma^2)^{d/2}} e^{-X^{2/d}/2\sigma^2}, 
    \label{eqn:PX}
\end{equation}
which in terms of $\log\mathcal{L}$ becomes
\begin{equation}
    \mathcal{P}(\log\mathcal{L}) = \frac{1}{\Gamma\left(\frac{d}{2}\right)}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
\end{equation}
i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, with mean and variance
\begin{equation}
    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
Given the prior is uniform $X\sim U(0,1)$, the KL divergence in the $\sigma^d \ll 1$ limit is:
\begin{equation}
\mathcal{D}_\mathrm{KL} = -\frac{d}{2}\log 2 \Gamma(1+\tfrac{d}{2})^{2/d} \sigma^2.
\end{equation}

@@

\subsection{The termination prior volume}\label{sec:termination}
Making the above discussion more precise, we wish to find the compression factor $\log X_ \mathrm{f}$ at which the termination criterion is met, which is larger in magnitude than $\DKL$ (\cref{fig:logX_distribution}). The difficulty is that at an intermediate iteration we only know the posterior up to the maximum log-likelihood live point, which until just before the end is far from the posterior bulk. 
\par
In order to get an idea of where the true posterior bulk sits, we need to predict what the posterior looks like past the highest live point. We do this by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\Like(X)$ traced out by the live and dead points. 
One would never use this predicted posterior to perform inference, since more accuracy can always be achieved by simply finishing the run. However, we will demonstrate it is sufficient for making a run-time prediction for $\log X_\mathrm{f}$. 
\par
Quantitatively, this proceeds as follows: fit a function $f(X, \phi)$  with some parameters $\phi$ to the known likelihood profile, which allows us to express the prior volume we need to compress to as
\begin{equation}
	\Delta \mathcal{Z} = \epsilon \mathcal{Z}_\mathrm{tot},
\end{equation}
or equivalently
\begin{equation}\label{endpoint}
	\int_0^{X_\mathrm{f}} f(X, \phi)\ \mathrm{d}X = \epsilon \left( \int_0^{X_i} f(X, \phi)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead} \right),
\end{equation}
where $X_i$ is the volume of the iteration we have currently compressed to, and $\mathcal{Z}_\mathrm{dead}$ is the evidence we have accumulated up to this point. $X_\mathrm{f}$ can then be identified by solving the above equation either analytically or numerically. 
\par
Once $X_\mathrm{f}$ is known, the corresponding iteration count depends on the live point schedule. For example, in the constant $\nlive$ case $\log X$ decreases by $1/n$ at each iteration, so the total number of iterations $N_\mathrm{f}$ would be
\begin{equation}
	N_\mathrm{f} = - n\log X_\mathrm{f} .
\end{equation}

@@

\subsection{Alternative approaches}
More comprehensive Bayesian approaches, perhaps including \textit{a priori} information about the likelihood or greater flexibility in the fitting function, could likely perform better than what we have just presented. However, such methods would not befit \textit{run-time} prediction which has a much more limited computational budget, hence the more pragmatic approach we have adopted. Here, we discuss as a benchmark alternative approaches to endpoint estimation that have a comparable computational complexity. 
\subsubsection{Integral progress}
An alternative approach used in \textsc{Ultranest} \citep{ultranest} derives a progress bar based on the fraction of the accumulated integral compared to the remaining integral, approximated as 
\begin{equation}\label{eq:integral_progress}
    \mathcal{Z}_\mathrm{rem} \approx \Like_\mathrm{max}^{\mathrm{live}} X_i.
\end{equation}
This has a several shortcomings. First, run-time is proportional to compression rather than accumulation of the integral, since it takes just as long to traverse the width of the bulk as it does any other width. Second, because of the point-like nature of the posterior mid-run, the remaining integral approximated as such holds nearly all of the evidence, so the relative fraction of the accumulated and remaining evidence is almost zero for most of the run. Finally, approximation \eqref{eq:integral_progress} is always an underestimate, because as previously found the maximum live point is generally nowhere near the true maximum. This approach can be useful in low dimensions when the live points are always near the maximum, but in general is less reliable.
\par
We also note that for implementations whose sampling efficiency $f_\mathrm{sampler}$ varies, the progress indicator of \cref{sec:termination} does not directly correspond to time, but will still be useful if the variation is known \textit{a priori}. For example, ellipsoidal samplers such as \textsc{MultiNest} and the MLFriends algorithm in \textsc{UltraNest} reduce in efficiency near the posterior bulk, so an impatient user who does not wish to expend the extra exponential effort for the final few percent can confidently halt earlier if they know they are already near the termination point.
